Approximation as Differentiation

### 2025-06-09 main/first-order.tex
Pared back version of biproducts.

For differentiable functions, we can combine linear approximations by adding tangent vectors. For stable
functions, we can combine approximants by taking their meet. The additive structure in both cases form a
commutative monoid with an addition + and zero element 0. A category $\cat{C}$ is enriched in $\CMon$, the
(monoidal) category of commutative monoids, when every hom-set $\cat{C}(X,Y)$ provides an operation + to add
morphisms together plus a zero morphism $0: X \to Y$, and composition is \emph{bilinear}, i.e.~given by a
family of morphisms $\Hom{\cat{C}}{Y}{Z} \tensor \Hom{\cat{C}}{X}{Y} \to \Hom{\cat{C}}{X}{Z}$ in $\CMon$ that
preserve the additive structure in $\Hom{\cat{C}}{Y}{Z}$ and $\Hom{\cat{C}}{X}{Y}$ separately:

@@

In any $\CMon$-category,

\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_X \comp \biinj_X &= \id_X \\
   \biproj_Y \comp \biinj_X &= \zero_{X,Y}
\end{salign*}
\end{center}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_Y \comp \biinj_Y &= \id_Y \\
   \biproj_X \comp \biinj_Y &= \zero_{Y,X}
\end{salign*}
\end{center}
\end{minipage}

@@

\hrule

We start with the idea of a biproduct and semi-additive categories, which are always enriched in $\CMon$, the
category of commutative monoids.

\begin{definition}[Semi-additive category]
\label{def:biproducts:semi-additive}
A category with finite products and coproducts is \emph{semi-additive} if the canonical morphisms (projections
and injections) give an isomorphism
\[\textstyle X \coprod Y \iso X \times Y\] that is natural in both variables.
\end{definition}

The product/coproduct is called a \emph{biproduct}, with the biproduct structure denoted by $(\biprod, 0)$ and
projections $\biproj_X, \biproj_Y$ and injections $\biinj_X$ and $\biinj_Y$. The unit $0$ is both terminal and
initial and is called a \emph{zero} object. A semi-additive category $\cat{C}$ is enriched in $\CMon$: for any
two morphisms $f, g: X \to Y$ in $\cat{C}$, the biproduct structure provides a way to ``add'' them together,
forming a morphism $f + g: X \to Y$. Diagrammatically:

\begin{center}
\begin{tikzcd}
   X \arrow[r, "\diag"] & X \biprod X \arrow[r, "f \biprod g"] & Y \biprod Y \arrow[r, "\codiag"] & Y
\end{tikzcd}
\end{center}

Here $\diag$ denotes the diagonal $\prodM{\id_X}{\id_X}$ given by the universal property of the product and
$\codiag$ denotes the codiagonal $\coprodM{\id_X}{\id_Y}$ given by the universal property of the coproduct.
The $f \oplus g$ morphism is the component-wise map $\prodM{f \comp \biproj_X}{g \comp \biproj_Y} =
\coprodM{\biinj_X \comp f}{\biinj_Y \comp g}$. Similarly we can exhibit a \emph{zero} morphism $0_{X,Y}$ by
composing the unique maps in and out of the zero object:

\begin{center}
\begin{tikzcd}
   X \arrow[r, "!_X"] & 0 \arrow[r, "!^Y"] & Y
\end{tikzcd}
\end{center}

It is easy to verify that $+$ is associative and commutative and that $f + 0_{X,Y} = f$, and thus that every
hom-object $\cat{C}(X,Y)$ is an object in $\CMon$. Moreover composition is \emph{bilinear}, i.e.~given by a
family of morphisms $\Hom{\cat{C}}{Y}{Z} \tensor \Hom{\cat{C}}{X}{Y} \to \Hom{\cat{C}}{X}{Z}$ in $\CMon$ that
preserve the additive structure in $\Hom{\cat{C}}{Y}{Z}$ and $\Hom{\cat{C}}{X}{Y}$ separately:

\begin{salign*}
f \comp \zero_{X,Y} = 0_{X,Z} = \zero_{Y,Z} \comp f
\end{salign*}
\begin{salign*}
(f + g) \comp h &= (f \comp h) + (g \comp h) \\
h \comp (f + g) &= (h \comp f) + (h \comp g)
\end{salign*}

\subsubsection{Biproduct laws}

It is also possible to use these laws to define biproducts, in the case where $\cat{C}$ is enriched
in $\CMon$ and so equipped with addition of morphisms and zero morphisms. In that case any products or
coproducts in $\cat{C}$ are necessarily biproducts.

\begin{proposition}
\label{prop:biproducts:from-product-or-coproduct}
In a category $\cat{C}$ enriched in $\CMon$:
\begin{enumerate}
\item If $(X \times Y, \pi_1, \pi_2)$ is a product then $(X \times Y, \inj_1, \inj_2)$ is a coproduct, where
$\inj_X = \prodM{\id_X}{0_{X,Y}}: X \to X \times Y$ and $\inj_Y = \prodM{0_{X,Y}}{\id_Y}: Y \to X \times Y$.
\item If $(X \coprod Y, \inj_1, \inj_2)$ is a coproduct then $(X \coprod Y, \pi_1, \pi_2)$ is a product, where
$\pi_1 = \coprodM{\id_X}{0_{Y,X}}: X \coprod Y \to X$ and $\pi_2 = \coprodM{0_{X,Y}}{\id_Y}: X \coprod Y \to
Y$.
\item If $X$ is terminal (resp.~initial) then $X$ is initial (resp.~terminal).
\end{enumerate}
\end{proposition}

A biproduct category $\cat{C}$ cannot be Cartesian closed without being trivial. If $0 \times X \iso X$ then
$\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X \Rightarrow Y} \iso 1$, and so $\cat{C}$ is terminal, with only a
single object up to isomorphism.

>>>>

For differentiable functions, we can combine linear approximations by adding tangent vectors; for stable
functions, we can combine approximants by taking their meet. In both cases the additive structure is given by
a commutative monoid with addition + and zero element 0. A category $\cat{C}$ is enriched in $\CMon$, the
(monoidal) category of commutative monoids, when every hom-set $\cat{C}(X,Y)$ provides an operation + to add
morphisms together plus a zero morphism $0: X \to Y$, and composition is \emph{bilinear}, i.e.~given by a
family of morphisms $\Hom{\cat{C}}{Y}{Z} \tensor \Hom{\cat{C}}{X}{Y} \to \Hom{\cat{C}}{X}{Z}$ in $\CMon$ that
preserve the additive structure in $\Hom{\cat{C}}{Y}{Z}$ and $\Hom{\cat{C}}{X}{Y}$ separately:

@@

In any $\CMon$-category, any products and coproducts coincide.

\begin{proposition}
\label{prop:biproducts:from-product-or-coproduct}
In any $\CMon$-category:
\begin{enumerate}
\item If $X \times Y$ is a product then it is also coproduct, with $\inj_X = \prodM{\id_X}{0_{X,Y}}$ and
$\inj_Y = \prodM{0_{X,Y}}{\id_Y}$.
\item If $X \coprod Y$ is a coproduct then it is also a product, with $\pi_1 = \coprodM{\id_X}{0_{Y,X}}$ and
$\pi_2 = \coprodM{0_{X,Y}}{\id_Y}$.
\item $X$ is terminal iff $X$ is initial.
\end{enumerate}
\end{proposition}

An object that is both terminal and initial is called a \emph{zero} object; an object that is both a product
and a coproduct is called a \emph{biproduct}. More directly, a biproduct is as an object $X \biprod Y$
together with morphisms

\begin{center}
\begin{tikzcd}
   X \arrow[r, "\biinj_X", shift left] &
   X \biprod Y \arrow[l, "\biproj_X", shift left] \arrow[r, "\biproj_Y"', shift right] &
   Y \arrow[l, "\biinj_Y"', shift right]
\end{tikzcd}
\end{center}

\noindent satisfying

\vspace{-3mm}
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_X \comp \biinj_X &= \id_X \\
   \biproj_Y \comp \biinj_X &= \zero_{X,Y}
\end{salign*}
\end{center}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_Y \comp \biinj_Y &= \id_Y \\
   \biproj_X \comp \biinj_Y &= \zero_{Y,X}
\end{salign*}
\end{center}
\end{minipage}

@@

\noindent from which the universal properties of the product and coproduct easily follow. A category $\cat{C}$
with finite biproducts is called \emph{semi-additive}, and cannot be Cartesian closed without being trivial.
Since $X \iso 0 \biprod X$, Cartesian closure implies $\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X \Rightarrow
Y} \iso 1$, and all objects are isomorphic.

% in any such category the canonical projections and injections give an isomorphism $\textstyle X \coprod Y
% \iso X \times Y$ that is natural in both variables
%
% A semi-additive category $\cat{C}$ is enriched in $\CMon$: for any two morphisms $f, g: X \to Y$ in
% $\cat{C}$, the biproduct structure provides a way to ``add'' them together, forming a morphism $f + g: X \to
% Y$. Diagrammatically:
%
% \begin{center} \begin{tikzcd} X \arrow[r, "\diag"] & X \biprod X \arrow[r, "f \biprod g"] & Y \biprod Y
% \arrow[r, "\codiag"] & Y \end{tikzcd} \end{center}
%
% Here $\diag$ denotes the diagonal $\prodM{\id_X}{\id_X}$ given by the universal property of the product and
% $\codiag$ denotes the codiagonal $\coprodM{\id_X}{\id_Y}$ given by the universal property of the coproduct.
% The $f \oplus g$ morphism is the component-wise map $\prodM{f \comp \biproj_X}{g \comp \biproj_Y} =
% \coprodM{\biinj_X \comp f}{\biinj_Y \comp g}$. Similarly we can exhibit a \emph{zero} morphism $0_{X,Y}$ by
% composing the unique maps in and out of the zero object:
%
% \begin{center} \begin{tikzcd} X \arrow[r, "!_X"] & 0 \arrow[r, "!^Y"] & Y \end{tikzcd} \end{center}
%
% It is easy to verify that $+$ is associative and commutative and that $f + 0_{X,Y} = f$, and thus that every
% hom-object $\cat{C}(X,Y)$ is an object in $\CMon$.
%
% \subsubsection{Biproduct laws}
%
% It is also possible to use these laws to define biproducts, in the case where $\cat{C}$ is enriched in
% $\CMon$ and so equipped with addition of morphisms and zero morphisms. In that case any products or
% coproducts in $\cat{C}$ are necessarily biproducts.
